Delivering Reliable Software in the European Scientific Arena: Evolution of Methodologies and New Trend Analysis

### 2018-02-08 paper-ieee-trans.tex
last changes

\IEEEPARstart{D}uring the last 15 years the provision of an European e-Infrastructure,
which supports an unified access to large-scale computing and intense data analysis
applications, has been driven by the needs of scientific communities as part of their
pan-European research collaborations. One of the well-known examples is the Worldwide 
Large Hadron Collider (LHC) Computing Grid infrastructure, powered 
by the particle physics scientific community, built up to satisfy the computing needs
of processing, storing and analysing the huge quantity of data coming from the Large 
Hadron Collider particle accelerator. A significant number of innovation projects and
technology transfer activities -- embraced by the 5th, 6th and 7th Framework Programmes,
and ultimately by the Horizon 2020 programme \cite{h2020} -- are being held to coordinate and 
develop distributed solutions that provide a seamless and transparent access to those
computing and data resources, available for the researchers.

Major developments were done throughout the years, mainly focused on providing new
functionalities, leaving aside at early times the adoption of software engineering methodologies.
The software delivered was then less tested, resulting in constant
rollbacks and patches applied to production systems. Thereby, the stability and
reliability of the e-Infrastructures was often compromised, affecting
negatively their exploitation by the scientific communities.

@@

Analyzing the European Commission (EC)-funded software projects in the last decade
we have observed a continuous increase of the prominence and robustness of
testing and validation procedures. One of many reasons is related to the
evolution of the software engineering methodologies (creation of the Agile
Manifesto \cite{agile-manifesto}, and rise of DevOps practices \cite{zhu}) together
with the parallel advancements in the information and communication technology
field (such as virtualization, the enabling technology for Cloud computing and 
operating--system--level virtualization known as containerization). All of which
complemented by the emergence of automation and event-response tools.

@@

The EC projects considered in this text are listed chronologically in Table
\ref{tab:eup}. In most cases, each project paved the way of the subsequent projects,
following a logic evolution in which the quality and reliability of software have
increasingly had a more significant role. Experiences gathered during the 
course of each project support this fact and, as it is shown in Table \ref{tab:feat},
latest projects are the most committed in terms of software reliability awareness.
Table \ref{tab:feat} summarizes also the topics addressed in the following sections, highlighting 
the commonalities and specific features on software reliability.

@@

The {\sl DataGrid} \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) 
had as main goal to develop the software to provide basic Grid functionality
and associated management tools for a large scale testbed for demonstration projects in three
specific areas of science, such as particle physics, bioinformatics and earth observation.
%was devoted to develop the first European infrastructure for Grid computing. 
The project brought together 21 academic and industry partners, from 15 different
countries. 
The project delivered its own software distribution, named EDG (EU {\sl DataGrid}), strongly 
based on Globus middleware services \cite{globus}. The project was organized in 12 workpages, from 
which 5 were devoted to software development coordinated by an 
Architecture Task Force, supervising the overall design and technical consistency 
of the developments.

@@

Agile methodologies were being introduced at the time
through the Agile manifesto \cite{agile-manifesto}, as such the project's
development design suffered from the methods, tools, techniques and best
practices emerging from the new discipline of software engineering
\cite{agile}. Nonetheless, the project focused on the experiences and procedures
used by different open source software projects, such as Linux and Apache, trying to reach a
higher maturity level \cite{cmm}.

@@

The source code was available at a private, centralized version control system.
The code was passed through a manual certification procedure at the time of the release. 
This procedure tried to improve the reliability of the software components by applying an
acceptance criteria plan at each stage of the software release process 
\cite{egee:acceptance-criteria} i.e. integration, certification, pre--production and 
production. Starting with {\sl EGEE--II}, the project adopted automation in the software
lifecycle process by leveraging the automatic build system for Grid middleware, that is
the {\sl ETICS} \cite{etics} solution.

@@

The {\sl EGI--InSPIRE} (Integrated Sustainable Pan--European Infrastructure for
Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire}
arose as the continuation of the {\sl EGEE--III} project, with the mission of establishing and
maintaining a sustainable European Grid Infrastructure.

{\sl EGI--InSPIRE} not only ensured the coordination of the operation of a secure, reliable 
European-wide production Grid infrastructure federated from National Grid Initiatives (NGIs)
that is integrated and interoperates with other Grids worldwide but also foreseed the 
maintenance and development of operational tools, like the Operations Portal \cite{egi-ops}, 
the {\sl EGI} Helpdesk \cite{ggus} or Grid configuration repository (GOCDB) \cite{gocdb} and the 
coordination of the external provision of software required by {\sl EGI} to 
form the production infrastructure. For the later it maintained a production--ready 
Grid computing middleware distribution called {\sl UMD} (Unified Middleware Distribution), and 
defined general and component specific quality criteria to be applied to software 
components made available through this repository.
The role of {\sl UMD} was to enforce the fulfilment of this set of quality
criteria \cite{egi-qc} in the software being delivered (i.e.,
{\sl EMI} and {\sl Globus}) \cite{mario}. {\sl UMD} is still being used and
deployed in the European scientific e--Infrastructures under a follow--up
project, the {\sl EGI--Engage} (Engaging the EGI Community towards an Open Science
Commons) \cite{cordis:egi-engage}. This distribution is currently complemented
by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution).

@@

In the following section we will describe the three main pillars that sustain the project's strategy to seek reliability throughout the 
software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
and metrics gathering definition} will provide guidelines to develop quality 
software, validating each change in the code, and facilitate its adoption. The software lifecycle is continuously monitored
by a set of metrics, allowing promptly reactions to detected malfunctions throughout the
process. A \textit{bottom-up adoption of DevOps principles} will promote automation to enable
the foregoing change-based development process and set up advanced delivery mechanisms.
The last key pillar is constituted by a \textit{two-factor validation} carried out over the project's pilot testbeds and
e-Infrastructure resource provider's early-adoption.

@@

\item \textit{Code review}.
Represents the last step in the change management pipeline, once the candidate change has 
successfully passed through the testing methods described previously. It implies the human-based
revision of the proposed change to discuss its adequance in terms of scope, objective fulfillment, 
documentation completeness, etc. On approval, the candidate change is definitely merged into
production. Human code reviewing is paramount in the software quality assessment, specially 
important when the SQA testing requirements are fully automated. Secure code reviews are
done at this stage, assessing common vulnerabilities from inputs coming from automated linters
and manual dynamic application security testing.

@@

\item \textit{Automated Deployment}.
In order to lower the barriers of software adoption, the SQA criteria requires the automated 
deployment of the products delivered as part of the catalogue. Automation in this context is 
tackled using a configuration management tool. These tools allow to have an 
\textit{infrastructure as code} (IaC), managing the component's deployment through declarative
definitions. The definition files appear as additional source of documentation, 
self-documenting code, as they sequentially guide the component's deployment process in multiple
platforms. In this regard, {\sl INDIGO--DataCloud} project contributed to open--source IaC tools such 
as Ansible \cite{indigo-ansible} and Puppet \cite{indigo-puppet}. A representative example of 
such contribution are the \textit{50 roles} being developed from scratch, currently hosted in
the Ansible Galaxy portal \cite{indigo-galaxy}. Figure 
\ref{fig:fig_confman} shows the number of products that offer an automated means for deployment.
It shows an increase after the {\sl INDIGO--1} release especially in the previous weeks of each
of the two major releases.
\end{itemize}

@@

@@

>>>>

\IEEEPARstart{D}uring the last 15 years the provision of an European e-Infrastructure,
which supports an unified access to large-scale computing and intense data analysis
applications, has been driven by the needs of scientific communities as part of their
pan-European research collaborations. One of the well-known examples is the Worldwide 
Large Hadron Collider (LHC) Computing Grid infrastructure, powered 
by the particle physics scientific community, built up to satisfy the computing needs
of processing, storing and analysing the huge quantity of data coming from the 
LHC particle accelerator. A significant number of innovation projects and
technology transfer activities -- embraced by the 5th, 6th and 7th Framework Programmes,
and ultimately by the Horizon 2020 programme \cite{h2020} -- are being held to coordinate and 
develop distributed solutions that provide a seamless and transparent access to those
computing and data resources, available for the researchers.

Major developments were done throughout the years, mainly focused on providing new
functionalities, leaving aside in the early times the adoption of software engineering methodologies.
The software delivered was then less tested, resulting in constant
rollbacks and patches applied to production systems. Thereby, the stability and
reliability of the e-Infrastructures was often compromised, affecting
negatively their exploitation by the scientific communities.

@@

By analyzing the European Commission (EC)-funded software projects in the last decade
we have observed a continuous increase of the prominence and robustness of
testing and validation procedures. One of many reasons is related to the
evolution of the software engineering methodologies (creation of the Agile
Manifesto \cite{agile-manifesto}, and rise of DevOps practices \cite{zhu}) together
with the parallel advancements in the information and communication technology
field (such as virtualization, the enabling technology for Cloud computing and 
operating--system--level virtualization known as containerization). All of which
complemented by the emergence of automation and event-response tools.

@@

The EC projects considered in this text are listed chronologically in Table
\ref{tab:eup}. In most cases, each project paved the way of the subsequent projects,
following a logic evolution in which the quality and reliability of software have
increasingly had a more significant role. Experiences gathered during the 
course of each project support this fact and, as it is shown in Table \ref{tab:feat},
latest projects are the most committed in terms of software reliability awareness.
Table \ref{tab:feat} summarizes also the topics addressed in the following sections, highlighting both
the commonalities and specific features on software reliability.

@@

The {\sl DataGrid} \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) 
had as the main goal to provide scientific communities, including physics, biology and 
earth sciences, with intensive computing and large-scale dataset analysis capabilites.
The project brought together 21 academic and industry partners, from 15 different countries \cite{gagliardi}. 
Grid technology was the emerging technology to be used in order to address these requirements,
thus the project delivered its own software distribution, named EDG (EU {\sl DataGrid}), strongly 
based on Globus middleware services \cite{globus}. The project was organized in 12 workpages, from 
which 5 were devoted to software development coordinated by an 
Architecture Task Force, supervising the overall design and technical consistency 
of the developments.

@@

The Agile manifesto \cite{agile-manifesto} was by then an emerging methodology, thus not
considered by the project, and as such the project's
development design suffered from the methods, tools, techniques and best
practices emerging from the new discipline of software engineering
\cite{agile}. Nonetheless, the project focused on the experiences and procedures
used by diverse collaborative open source software projects, such as Linux and Apache, trying to 
reach a higher maturity level \cite{cmm}.

@@

The source code was available at a private, centralized version control system.
The code passed through a manual certification procedure at the time of the release. 
This procedure tried to improve the reliability of the software components by applying an
acceptance criteria plan at each stage of the software release process 
\cite{egee:acceptance-criteria} i.e. integration, certification, pre--production and 
production. Starting with {\sl EGEE--II}, the project adopted automation in the software
lifecycle process by leveraging the automatic build system for Grid middleware, that is
the {\sl ETICS} \cite{etics} solution.

@@

The {\sl EGI--InSPIRE} (Integrated Sustainable Pan--European Infrastructure for
Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire}
arose as the continuation of the {\sl EGEE--III} project, with the mission of establishing and
maintaining a sustainable European Grid Infrastructure, composed by a federation of National 
Grid Initiatives and interoperable with other Grids worldwide.
To accomplish this goal the project foreseed the 
maintenance and development of operational tools, like the Operations Portal \cite{egi-ops}, 
the {\sl EGI} Helpdesk \cite{ggus} or the Grid configuration database (GOCDB) \cite{gocdb}, and the 
management of the software provisioning process \cite{mario}, to validate and distribute the software that conformed
the production infrastructure. As a result of this process, a production--ready 
Grid computing middleware distribution named {\sl UMD} (Unified Middleware Distribution) was
maintained, comprised by roughly 250 software components supplied by diverse technology providers such 
as {\sl EMI} and {\sl Globus}. In order to be distributed through UMD
repositories, the software has to be compliant with EGI's Quality Criteria (QC) \cite{egi-qc}. The 
QC defines a set of requirements in different areas: documentation, deployment, security, information
model and operations.
{\sl UMD} is still being used and
deployed in the European scientific e--Infrastructures under a follow--up
project, the {\sl EGI--Engage} (Engaging the EGI Community towards an Open Science
Commons) \cite{cordis:egi-engage}. This distribution is currently complemented
by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution).

@@

In the following sections we will describe the three main pillars that sustain the project's strategy to seek reliability throughout the 
software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
and metrics gathering definition} will provide guidelines to develop quality 
software, validating each change in the code, and facilitate its adoption. The software lifecycle is continuously monitored
by a set of metrics, allowing promptly reactions to detected malfunctions throughout the
process. A \textit{bottom-up adoption of DevOps principles} will promote automation to enable
the foregoing change-based development process and set up advanced delivery mechanisms.
The last key pillar is constituted by a \textit{two-factor validation} carried out over the project's pilot testbeds and
e-Infrastructure resource provider's early-adoption.

@@

\item \textit{Code review}.
Represents the last step in the change management pipeline, once the candidate change has 
successfully passed through the testing methods described previously. It implies the human-based
revision of the proposed change to discuss its adequance in terms of scope, objective fulfillment, 
documentation completeness, etc. On approval, the candidate change is definitely merged into
the source code's production version. Human code reviewing is paramount in the software quality assessment, specially 
important when the SQA testing requirements are fully automated. Secure code reviews are
done at this stage, assessing common vulnerabilities from inputs coming from automated linters
and manual dynamic application security testing.

@@

\item \textit{Automated Deployment}.
In order to lower the barriers of software adoption, the SQA criteria requires the automated 
deployment of the products delivered as part of the catalogue. Automation in this context is 
tackled using a configuration management tool. These tools allow to have an 
\textit{Infrastructure as Code} (IaC), managing the component's deployment through declarative
definitions. The definition files appear as additional source of documentation
-- self-documenting code -- as they sequentially guide the component's deployment process in multiple
platforms. In this regard, {\sl INDIGO--DataCloud} project contributed to open--source IaC tools such 
as Ansible \cite{indigo-ansible} and Puppet \cite{indigo-puppet}. A representative example of 
such contribution are the \textit{50 roles} being developed from scratch, currently hosted in
the Ansible Galaxy portal \cite{indigo-galaxy}. Figure 
\ref{fig:fig_confman} shows the number of products that offer an automated means for deployment.
It shows an increase after the {\sl INDIGO--1} release especially in the previous weeks of each
of the two major releases.
\end{itemize}

@@

\bibitem{gagliardi}
Gagliardi F., Jones B., Reale M., Burke S., \emph{European DataGrid Project: Experiences of Deploying a Large Scale Testbed for E-science Applications},
in Calzarossa M.C., Tucci S. (eds) Performance Evaluation of Complex Systems: Techniques and Tools, Lecture Notes in Computer Science, vol 2459. Springer, 2002.

@@

\bibitem{egi-qc}
\emph{EGI Quality Criteria} [Online]. Available: 
https://egi-qc.github.io/
